Noisy Work: A Review of The Uncertainty Visualisation Literature

### 2024-02-07 paper.qmd
added aside comments in paper, also explanation for mathematical definition of uncertainty, also updated gitignore

# 1. Introduction
From entertainment choices to news articles to insurance plans, the modern citizen is so over run with information in every aspect of their life it can be overwhelming. So overwhelming in fact, that it is not uncommon for people to become so overwhelmed by the information they refuse to engage with it at all. In this overflow of information, tools that can effectively summarise information down into simple and clear ideas become more valuable. While websites that summarise insurance plans or hotel options down into a clear table make a living off it, few tools are as powerful or effective as a visualisation. Information visualisations remain one of the most powerful tools for fast and reliable science communication. Effectively leveraging visualisations to make them more effective is therefore key in facilitating simple and clear communication of science and other ideas.

@@

# 2. Defining Uncertainty
Uncertainty visualisation is not made any easier by the fact that the term "uncertainty" lacks a commonly accepted definition in the literature. @Spiegelhalter2017 even commented that the definition of uncertainty essentially "depends on who you ask". The disarray of organisation and definitions within the uncertainty literature is a common sentiment expressed by reviewers[@Spiegelhalter2017; @Kinkeldey2014]. This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is. In this section we will establish exactly what we mean when we talk about uncertainty.

## 2.1 Coloquial Definitions
The most common definition of uncertainty used in papers that discuss visualising it are colloquial. Most definitions make some nod towards variance, precision, trust, error, missing values or another related concepts. Since *uncertainty*, unlike *probability*, does not have a strict definition that is established in mathematics, there is large variety in its usage. Some consider it to be synonymous with probabiltiy, some consider it to be an encompassing umbrella term of which probability is only an element, and some consider it to by something else entirely. This lack of a definition leads to uncertainty being swapped out freely with other vague words such as ambiguity and risk. Definitions presented in papers that discuss uncertainty visualisation want it to be encompassing and include everything a layperson might consider when trying to consider "uncertainty". The issue with these broad definitions is is that in order to visualize something, it needs to be quantified, and in order to quantify something, we need to know exactly what it is we are counting. This discrepancy appears regularly in the uncertainty visualisation literature. Often a vague definition such as "uncertainty is anything that isn't deterministic", or even a refusal to define uncertainty at all in an experiments introduction will be accompanied by a highly specific quantification of uncertainty as a probability density distribution of a forecast when providing the visualisation. Ironically it seems that many of the papers that refer to themselves as "uncertainty" papers are just suffering from a form of uncertainty, a lack of precision.[^1]

@@

## 2.2 Taxonomy Definitions
### 2.2.1 Taxonomies of Uncertainty
While colloquial definitions of uncertainty are useful when trying to communicate with laypeople who don't understand strict statistical definitions, these definitions are not helpful when we are trying to work out exactly what should be shown on a plot. When authors try to make a considerable effort to define what we mean when we talk about uncertainty, they often express it in the form of a taxonomy.

@@

@@

@@

@@

@@

@@

# 3.General Attitude Towards Uncertainty (Uncertainty Communication Boradly)
It would be impossible to discuss uncertainty visualisation without mentioning the broader conversation around uncertainty communication. The conversation around uncertainty communication is often centered around the psychology is communicating risks.

Most research into general uncertainty communication seems to focus specifically on how we communicate risks. Risk and uncertainty are not perfect synonyms of one another. @Spiegelhalter2017 makes the distinction clear by specifying *risk* as *a precise random event* (such as a coin flip) while *uncertainty* (also sometimes referred to as abiguity) is imprecision (such as the estimated probability associated with a coin that might be biased). For this reason, risk is often seen as a specific case of uncertainty and that well defined sub area has a lot of detailled research. Some markers of probability that have common uses, probabilities such as 0 (impossible), 0.5 (a fair coin flip), 1 (certain) are easy for people to have an internal sense of [@Spiegelhalter2017]. It is incredibly hard to communicate small risks, because people cannot differentiate between 1/100 an 1/1000, and communicating low probability but high impact risks are particularly challenging [@Spiegelhalter2017]. This issue can be somewhat alleviated by communicating the likelihood of an event in relation to another event on a similar scale (such as being struck by lightning) [@Spiegelhalter2017]. The framing of how we communicate risks matters due to the affect heuristic, so there will be a difference in peoples behaviours if, for example, you tell people they have a 90% chance of survival vs a 10% chance of death, and when expressing a frequency, a larger numerator communicates a larger risk [@Spiegelhalter2017]. The reference class of a particular risk should be explicitly stated to avoid confusion, for example if we are communicating the chance of rain, the time period over which we are expecting this chance of rain (i.e. 50% chance today or this week) needs to be specified [@Spiegelhalter2017]. The problem with a risk focus is that it is only a small piece of the uncertainty pie, and very few of these findings can be reasonably assumed to go beyond risk communication. Even less can be assumed to translate to uncertainty visualisation.

@@

@@

@@

## 4.1 Similar work
There are a wealth of uncertainty communication and visualisation literature reviews that already exist.

@@

### A navigating a forest analogy
When communicating uncertainty information, it is important to ask if we should consider the bias to be variance. We often talk about the "bias variance trade off" but we rarely discuss bias as "cashed in" variance, which in some sense, it is. If you are struggling to understand how bias is just traded off variance, let us walk through an analogy.

Let us consider the data analysis pipeline a dense forest that you need to cross to get to your house. To prevent yourself from getting *too* lost you decide to bring some poles and rope so you can mark where you are going. You know your rope is long enough to get you to your house with enough wiggle room that you can reach the 5 houses either side of your house. This problem is drawn in @fig-forest1. @fig-forest2 depicts the typical data analysis pipeline and resulting uncertainty using this forest analogy. In this method, we navigate the forest by putting one of our poles down every time we notice the trees have changed (i.e. we have move to a new section of the forest). As we move down the line and we put our poles down, we cannot be certain we are always walking in the direction of our house, but we can do small checks or validation to make sure we are vaguely heading in the right direction. When we get to the final stage, we have enough rope remaining to get to the three houses directly in front of us, and we state that band as our "uncertainty". @fig-forest3 depicts the bias variance trade off in the case where we make no decisions, record as much information as possible, and check every single model with every single variable and keep every alternative as a possible option. This method will result in no bias (although it is often impossible to check the entirety of these spaces so it cannot) but a huge amount of variance.  In an effort to avoid putting down any poles, we have managed to navigate the forest, but have essentially no idea which of the 10 houses are ours. Of course, with sufficient feedback we can whittle down this vast array of choices (which is what a neural network can do) but feedback from an increased sample size can only influence variance, it cannot influence bias that was previously variance. @fig-forest4 shows an alternative way to perform the data analysis process to consider variance at stages beyond model uncertainty (such as the scenario uncertainty that comes with different human choices in climate change models). In this scenario you navigate the forest to a certain point, and then clone yourself three times and each put down a new pole at different distances to each other. All four versions of yourself will then walk in an identical fashion from the pole to the end of the forest, and each have your own path with your own confidence interval. This is how the [climate scenario uncertainty](https://climatechange.chicago.gov/sites/production/files/2016-07/scenariotempgraph_0.jpg) is constructed.

This analogy may seem silly but it is explaining a very real problem that exists in uncertainty communication. The bias and variance trade off is commonly discussed in modelling, but rarely mentioned when we discuss communicating uncertainty. The reality is that every point in our data analysis, is full of decisions. Every time we make a decision at some point in this pipeline, we are impacting this bias and variance trade off. We have swapped out the variance from a large range of choices for the bias that may arise from an incorrect decision. Validation methods (such as cross validation, assumption checks, etc) help to guide the choices we make in the data analysis pipeline, but they can never offer any kind of certainty or truth considering validation methods themselves are also only estimates. When we consider the "uncertainty" of a model we usually only consider the variance of the final stage, the prediction error for a specific model and a specific set of features, and ignore the variance (that has been converted into bias by our selections) that was introduced at earlier stages (e.g. such as in data collection, when cleaning the data, or when considering alternative models). This is *well known* in modelling, it is why big tech companies try to collect as much data as they can and then put it in a neural network and hope they have enough data and enough feedback to avoid the this bias at *all* stages of the analysis. Recording *everything* means we don't even have to put any poles down at the data collection stage. Sometimes these statements are couched in a single assumption, such as when a time series model's confidence interval is constructed "assuming previous trends will continue into the future". This trade off issue means the uncertainty we typically estimate, in the form of a confidence inteval or traces or whatever you choose, has confidence intervals that are too small for what you actually consider the goal of the analysis to be. Unfortunately factors that typically decrease the variance, such as sample size, can only have an effect at the levels where that information is used to inform the choices. So confidence intervals that are incredibly small but built on a biased sample will be misleading and will be too small to *actually* predict future data. It is not uncommon for true results or models produced by a different team to sit entirely outside of the confidence band of another. This might be obvious when it is read, but the conversation about whether or not we should consider what was previously variance, to be variance when we communicate our analysis, is something that is rarely discussed although it is important.

Therefore, we must ask ourselves, *why* are we visualsing uncertainty? Uncertainty is only thought of as the final stages because to give a point prediction for a random process feels silly. By hiding the selections we made at earlier stages, when we traded that variance in for bias, we functionally made *point predictions* but it was just invisible to us.

@@

## 4.1?? In other statistics
What is meant by "uncertainty" may seem obvious to some, but when you attempt to quantify or visualise it you will quickly find yourself asking, "uncertainty about... what?". Do you mean uncertainty on an estimate? On a forecast? How many steps ahead is this forecast? Are we only considering the uncertainty in the estimate or in the parameters or are we considering the possibility of measurement error or biased inputs? Signal and noise can only be untangled in the presence of a motivating question.

>>>>

# 1. Introduction
::: aside
I have not touched this section since the confirmation so it would be best to ignore it
:::

From entertainment choices to news articles to insurance plans, the modern citizen is so over run with information in every aspect of their life it can be overwhelming. So overwhelming in fact, that it is not uncommon for people to become so overwhelmed by the information they refuse to engage with it at all. In this overflow of information, tools that can effectively summarise information down into simple and clear ideas become more valuable. While websites that summarise insurance plans or hotel options down into a clear table make a living off it, few tools are as powerful or effective as a visualisation. Information visualisations remain one of the most powerful tools for fast and reliable science communication. Effectively leveraging visualisations to make them more effective is therefore key in facilitating simple and clear communication of science and other ideas.

@@

# 2. Defining Uncertainty
::: aside
I have thought more about the idea of defining uncertainty mathematically and I actually *DO* think it is possible. I want the record to show the reason I *did not* do this earlier is that *nobody* in the uncertainty literature has defined uncertainty mathematically, and I thought I was supposed to highlight problems in the literature review, not solve them. I have had an idea of how to formalise uncertainty after seeing Xiao-li Meng's work, but an uncertainty formalisation would solve *several* open questions he has postulated in previous papers, and would be so new it would need to be a paper in of itself, not as a sub-section in a literature review. A mathematical definition would be *completely* new and *very complicated*, and I think its difficulty is pretty clear by the fact that *nobody* and *I really mean nobody*, not spiegelhatter or wilkinson or mceachren or any statistician working in this space, has provided an overarching mathematical definition. This is work that would take a long time (and it is the reason I started reading philosophy books because I was actually trying to do this) and even if I did do it, it *should not* be in this paper. I think it might be possible based on the reading I have been doing over the past year to try and formalise this work, *but it is not a quick thing that can just be added to this paper*. The reason I wanted to do a summary of the papers using the larger taxonomy of uncertainty was because it is encompasing of all the taxonomies, and it allows me to show  that there are *no* formal rules to the uncertainty that is used in uncertainty communication and visualisation. Here I am trying to make the point that the lack of a formal definition or of mathematical considerations *has created problems in the literature*. To clearly and mathematically define uncertainty will *be solving this problem*. Something I was told *not to do* because this is a literature review. If a mathematical definition that we are all happy with *needs* to be in this paper, then there is a reasonable chance this paper will never be finished.
:::

Uncertainty visualisation is not made any easier by the fact that the term "uncertainty" lacks a commonly accepted definition in the literature. @Spiegelhalter2017 even commented that the definition of uncertainty essentially "depends on who you ask". The disarray of organisation and definitions within the uncertainty literature is a common sentiment expressed by reviewers[@Spiegelhalter2017; @Kinkeldey2014]. This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is. In this section we will establish exactly what we mean when we talk about uncertainty.

## 2.1 Coloquial Definitions
::: aside
I am not sure how to collect these together and show that they are highly varied. This might be a situation for the tidytext clustering or something. There is an additional problem is that the definition for uncertainty is *completely* unrelated to the uncertainty that is quantified in the graphics. I think the tidytex can help with the definitions being all over the place, but I am not sure how to identify the lack of correlation between the colloquial definitions of uncertainty and the mathematical quantificaitons.
:::

The most common definition of uncertainty used in papers that discuss visualising it are colloquial. Most definitions make some nod towards variance, precision, trust, error, missing values or another related concepts. Since *uncertainty*, unlike *probability*, does not have a strict definition that is established in mathematics, there is large variety in its usage. Some consider it to be synonymous with probabiltiy, some consider it to be an encompassing umbrella term of which probability is only an element, and some consider it to by something else entirely. This lack of a definition leads to uncertainty being swapped out freely with other vague words such as ambiguity and risk. Definitions presented in papers that discuss uncertainty visualisation want it to be encompassing and include everything a layperson might consider when trying to consider "uncertainty". The issue with these broad definitions is is that in order to visualize something, it needs to be quantified, and in order to quantify something, we need to know exactly what it is we are counting. This discrepancy appears regularly in the uncertainty visualisation literature. Often a vague definition such as "uncertainty is anything that isn't deterministic", or even a refusal to define uncertainty at all in an experiments introduction will be accompanied by a highly specific quantification of uncertainty as a probability density distribution of a forecast when providing the visualisation. Ironically it seems that many of the papers that refer to themselves as "uncertainty" papers are just suffering from a form of uncertainty, a lack of precision.[^1]

@@

## 2.2 Taxonomy Definitions
### 2.2.1 Taxonomies of Uncertainty
::: aside
I actually think I agree with the sentiment in the grammar of graphics that a taxonomy is more of an opinion and we can reclassify it and is therefore not thorough enough to remain unchanged and be build upon. That being said, if this is a literature review and all existing literature use written definitions or taxnomies, that is what I have to work with. The reason I selected the larger taxonomy is because the other taxonomies all *fit within it* and it can be used to define a space that shows the definitions are inconsistent. 
:::

While colloquial definitions of uncertainty are useful when trying to communicate with laypeople who don't understand strict statistical definitions, these definitions are not helpful when we are trying to work out exactly what should be shown on a plot. When authors try to make a considerable effort to define what we mean when we talk about uncertainty, they often express it in the form of a taxonomy.

@@

::: aside
This figure is not going to be in the final version of the paper. It is here as a placeholder. Ideally, I would replace with with a graphic that shows where all these taxonomies land in the location, level, nature taxonomy (since it is encompassing) and show they are very inconsistent and a new definition is pulled out for almost every paper (even in statistics papers)
:::

@@

::: aside
I had this here to explain why these are the considerations people make in statistics and how they relate to elements of the taxonomy to try and provide some justification, but If I am just going to show the definitions are inconsitent, I might get rid of this section. I also largely included it so you guys can see what the taxonomies are considering
:::

@@

#### Context comments
::: aside
I have this here to explain why context based taxonomies can be ignored when we are considering uncertainty. I did not go into too much detail because it would involve discussing visualisation which I am trying to avoid at this point. I actually think this quote from the grammar of graphics is exactly what I was trying to say. *"The system in this book is not a model for a GIS, however, because geography and statistics differ in a crucial respect. Geography is anchored in real space-time and statistics in abstract dimensions. This is a distinction along a continuum rather than a sharp break; after all, there is a whole field called spatial statistics (Cressie, 1991). But this difference in focus clearly means that a system optimized to handle geography will not be graceful when dealing with statistical graphics"*. Basically, my point is this section is still discussing the definitions of unertainty, we have not gotten to disussing visualisation so we are still in the abstract *statistics* domain. The only parts of spatial statistics that we care about in this domain are the statistical properties that come from spatial information, and *none* of those statistical properties generate uncertainty. This is not unique to the properties of spatial information, other statistics information that is important for modelling and assumptions, such as non-linearity, class levels, etc do not contribute to uncertainty. Operations on data, inherrent randomness, and missing information generate uncertainty, which is what we are talking about here. A confidence interval is a confidence interval whether it came from spatial data, temporal, or cross sectional data. *This does not mean I am saying spatial and temporal information is not valuable ever*, I am just saying it is not valuable *yet*. This information will be *highly relevant* when we discuss context and translating from the abstract domain back into real world for communication. Since we are still in the abstract domain, things that are anchored in real space-time create confusion that is evident in the literature. The confusion that is created in the literature was going to be the topic I had later. Spatial and temporal uncertainty papers limit how they communicate their uncertainty because they are considering what can be visualised, not what they should *try* to visualise based on the uncertainty abstraction. Since the context is only relevant *with respect to uncertainty* when we visualise it, I was going to discuss this issue in the visualisation literature review section.
:::

@@

#### Maths comments I might relate back to an actual mathematical definition
::: aside
 In the "semantics" section of the book, Leeland Wilkinson provides a series of definitions of words related to uncertainty and mathematical definitions of examples of uncertainty, but those are not mathematical definitions of uncertainty. Every *mathematical* definition of uncertainty is a definition of *probability* OR a definition of an *example* of uncertainty, such as a confidence interval or PDF. I included this here, however, I actually now believe there actually is a mathematical definition, but it would not fit in this paper. If this paper remains a *here are the discussions of uncertainty and there is no mathematical definition* I might leave this here as an explanation as to why, but not suggest that there *isnt* a mathematical definition.
:::

@@

#### An explanation of the building blocks for my mathematical definition
::: aside
I wrote this section between me deciding it would be possible to mathematically define uncertainty and me deciding it would be too large for this paper. I am aware that this is not a formal definition and I cannot say this is the unifying feature of uncertainty without some proof. But, this is something that *can be* defined in mathematical terms and *can* lead to an encompassing definition of uncertainty (i.e. it would be something that the groups of confidence intervals, error bars, and missing data using an information filtration) but that is beyond the scope of this paper. This way of thinking about uncertainty has resolved a lot of my issues with the field and a lot of other issues that have been raised by authors in their summary of the literature. This is very similar to Xiao-li Meng's work that asserts that bias and variance are actually the same thing, with a couple of changes.
:::

The first problem in the conception and attempts to offer a definition of uncertainty, is that bias *and* variance are both technically uncertainty.

Basically the idea is that every analysis has some goal, and every choice we make that cuts out several alternatives is a "restriction" that trades the variance of multiple options into a potential bias. This occurs at *every stage* of the data analysis pipeline, from collecting data to communication. Bias comes from "incorrect information" being passed along the pipeline, while variance comes from "holding on to information". The randomness that is "irriduciable error" (if it exists or not is not relevant to our definition) cannot be differentiated from the bias and variance (I say both bias and variance because at the end of the pipeline it is impossible to know if error is being caused by bias or variance) unless we literally know the function that generated the data. We typically talk about bias and variance in terms of the model, but it exists at every stage of the pipeline. Since every step of the analysis pipeline has an almost infinite set of choices (there is an infinite set of things we can observe, an infinite number of choices that can be made in data cleaning, etc) each of these levels is functionally its *own* random variable. When you make your choice you completely shrink the variance of that level to 0 (it basically goes from a distribution of choices to a single outcome of that distribution) and, the amount this *guess* is wrong by, is bias.

This idea is very similar to the bias-variance trade off, but it is slightly different because we are assuming there is not a trade-off perse, but that bias is just "cashed in" variance. If selections at each stage are made using feedback, we can reduce the bias without increasing the variance, but it "costs" some data. Assumptions let you reduce the variance with a low data "cost" but there is a higher chance of being incorrect (since there is less validation) and introducing bias. Flexible models like neural networks can reduce the variance but it comes at a high data "cost". If you have a *lot* of data it is better to avoid as many assumptions as possible because increased certainty that comes from more data can only effect error that is still in the form of variance (i.e. we are considering alternatives). This is well established in modelling, it is why big tech companies try to collect as much data as they can and then put it in a neural network and hope they have enough data and enough feedback to avoid the bias at *all* stages of the analysis. Recording *everything* and putting it in the most flexible modle possible means we don't even have to put any poles down, even at the data collection stage

Validation methods (such as cross validation, assumption checks, etc) help to guide the choices we make in the data analysis pipeline, but they can never offer any kind of certainty or truth considering validation methods themselves are also only estimates. Once a variance has been cashed in for bias, its error cannot be reduced by more data.

When we consider the "uncertainty" of a model we usually only consider the variance of the final stage, the prediction error for *a specific model and a specific set of features*, and ignore the variance that has been converted into bias by our selections that was introduced at earlier stages of the pipeline. Since uncertainty is basically "how true is this estimate or statement", if we had not cashed in our variance for a single outcome of bias, it would appear as variance. The climate scenario uncertainty plots are an example of bias that was converted back to variance for the sake of the visualisation (time series forecasts typically assume away the bias by stating "we are assuming previous trends will continue into the future"). Prediction error is a useful metric because it allows us to see if the bias introduced in earlier stages expressed as variance through an error, however it still cannot account for bias introduced at the sampling stage as that bias also exists in the test set.

Prediction error is not always feasible or useful in a statistical analysis (such as when we produce a visualisation) so it is unclear exactly *how* we should reintroduce that bias as variance to communicate the pair as uncertainty. It is not uncommon for uncertainty to only be considered at the final stage of an analysis as a confidence interval or estimate range, but this is typically misleading as earlier bias has not been converted back to variance. Factors that typically decrease the variance, such as sample size, can only have an effect on something that is still in a state of variance. This means that an established model that is build on more and more data will have confidence intervals that are incredibly small but if it is build on some bias (such as a biased sample) the confidence intervals will become misleading and will be too small to *actually* predict future data. It is not uncommon for true results or models produced by a different teams to sit entirely outside of the confidence band of another. By hiding the selections we made at earlier stages, when we traded that variance in for bias, we functionally made *point predictions* (they were based on educated guess but point predictions none the less) but *these* point predictions are often invisible to us.

#### A Forest analogy
Let us consider the data analysis pipeline a dense forest that you need to cross to get to your house. To prevent yourself from getting *too* lost you decide to bring some poles and rope so you can mark where you are going. You know your rope is long enough to get you to your house with enough wiggle room that you can reach the 5 houses either side of your house. This problem is drawn in @fig-forest1. @fig-forest2 depicts the typical data analysis pipeline and resulting uncertainty using this forest analogy. In this method, we navigate the forest by putting one of our poles down every time we notice the trees have changed (i.e. we have move to a new section of the forest). As we move down the line and we put our poles down, we cannot be certain we are always walking in the direction of our house, but we can do small checks or validation to make sure we are vaguely heading in the right direction. When we get to the final stage, we have enough rope remaining to get to the three houses directly in front of us, and we state that band as our "uncertainty". @fig-forest3 depicts the bias variance trade off in the case where we make no decisions, record as much information as possible, and check every single model with every single variable and keep every alternative as a possible option. This method will result in no bias (although it is often impossible to check the entirety of these spaces so it cannot) but a huge amount of variance.  In an effort to avoid putting down any poles, we have managed to navigate the forest, but have essentially no idea which of the 10 houses are ours. Of course, with sufficient feedback we can whittle down this vast array of choices (which is what a neural network can do) but feedback from an increased sample size can only influence variance, it cannot influence bias that was previously variance. @fig-forest4 shows an alternative way to perform the data analysis process to consider variance at stages beyond model uncertainty (such as the scenario uncertainty that comes with different human choices in climate change models). In this scenario you navigate the forest to a certain point, and then clone yourself three times and each put down a new pole at different distances to each other. All four versions of yourself will then walk in an identical fashion from the pole to the end of the forest, and each have your own path with your own confidence interval. This is how the [climate scenario uncertainty](https://climatechange.chicago.gov/sites/production/files/2016-07/scenariotempgraph_0.jpg) is constructed.

#### The influence of the "goal" of an analysis
Prediction error is also useful because the *goal* of our analysis is clearly defined in a metric. This is not always the case. When you attempt to quantify or visualise it you will quickly find yourself asking, "uncertainty about... what?". Do you mean uncertainty on an estimate? On a forecast? How many steps ahead is this forecast? Are we only considering the uncertainty in the estimate or in the parameters or are we considering the possibility of measurement error or biased inputs? These questions start to highlight another invisible problem in uncertainty definitions, signal and noise can only be untangled in the presence of a motivating question.

The idea that uncertainty can only be defined in the presence of a motivating question is well grounded in most areas of statistics. The entire process of data analysis, from deciding what should be observed as data through to communicating that data in a plot is governed by human decision and the goal of an analysis. At the philosophical level, applied statistics is simply taking real world entities and boiling them down into probabilistic objects, an ontological process that is largely dependent on our goals [@Otsuka2023]. When we move onto data provenance the issue persists, as what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the resolution of the question [@Meng2014]. After moving onto modelling this issue continues as each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to signal devoid of meaning [@Carlin2023]. Even at the final stages of visualisation a lack of understanding of the motivating question make it difficult to untangle what is signal and what is noise, leaving many uncertainty visualisation studies with conflicting results [@Kinkeldey2014]. These cases highlight that uncertainty is defined at *every* stage in relation to our motivating question. Discussions of uncertainty cannot be had if we are not clear *what* we are uncertain about. Once it is established what we are uncertainty about, we can consider the other elements of uncertainty that need to be defined.

This concept is a little harder to define mathematically, but we *can* consider that every analysis has some metric that is the target "goal" of the analysis. Refusing to define this target goal from the begining does not mean you avoid this problem, it just increases the bias and variance issues described earlier, because there is no way to check assumptions or correct the variance with data. This is actually why I was thinking about the concept of *suffient visual statistics*. If a graphic has a piece of information mapped to it (which we should know because we made the graphic ourselves) then that information can be extracted with a percecptual task and the graph has a *sufficient visual statistic*. I also think the idea of sufficient statistics is useful because it establishes that we can only know if we have extracted all the relevant information from our *if a metric we are collecting data for has been defined*. If a piece of information cannot be extracted with a perceptual task, it can only be found with assumptions and we introduce a bias at the communication stage. If the information does exist but it reuires a heuristic or some collection of perceptual tasks that might be harder for some people than others, we have maintained variance at the communication stage. A complicated visualisation with a *huge* amount of information can technically be used to answer a lot of questions, but the mental overload would create variance. This is why it is important to have your *goal* in mind when designing a visualisation and not assume it can or should be able to do everything. I suspect we can work out retroactively which visualisation contain which visual statistics, by systematically working our way through the grammar of graphics and then information that is contained in simple perceptual tasks.

Even if this cannot be used to give a mathematical definition of uncertainty, it is still soemthing that *should* be considered. This might be obvious when it is read, but the conversation about whether or not we should consider what was previously variance to be variance when we communicate our analysis, or if avoiding defining a *goal* metric can have detrimental effects on our communication are things that are rarely discussed although it is important.

::: aside
Di said this has already been shown and I believe it, but then I don't understand why the uncertainty visualisation literature seems to ignore it. This is what I was trying to show with the plot and question section of the excel spreadsheet. I was trying to highlight that when you ask participants to answer a question, to mark them as correct or incorrect you need to define a metric, and if the plot was created using a pipeline that *ignores* that metric, it will be passing on incorrect information and lead to mistakes when it is being read. Expecting any plot to be able to answer any question that has the basis of some random metric simply because the metric is related to uncertainty and so is the plot seems ridiculous to me, but its harder to find a paper that *doesn't* do this than a paper that *does*.
:::

@@

# 3.General Attitude Towards Uncertainty (Uncertainty Communication Boradly)
::: aside
This section is supposed to be an explanation of the general literature, but the definition problem makes it hard to sort the general literature. While there is that overview from spiegelhatter, it is mostly a discussion on risk communication, communicating ambiguous information is not discussed much. The focus on risk (uncertainty as signal) and not uncertainty (uncertainty as noise) is an issue, but the bigger problem is caused by the vagueness of uncertainty issue. Since a paper will be titled "uncertainty communication" it could be any number of things mathematically, but 90% of the time is it just communicating risks or uncertainty as signal (and very rarely uncertainty as noise). I can only tell which one a paper is *once I am in the method section* and even then it is sometimes not explained. Also, since I haven't been using this metric since the beginning I need to go back and check this for all the papers. I also think this section would benefit from the table, I can quantify how many papers communicate uncertaitny as signal, consider risk-aversion and truth but dont mention transparency, etc. I think the only way to back up a lot of my general comments is with these tables that show how widespread these issues are. This process is tedious but because definitions for key words are redefined within every paper, I dont think a clustering method will help
:::

It would be impossible to discuss uncertainty visualisation without mentioning the broader conversation around uncertainty communication. The conversation around uncertainty communication is often centered around the psychology is communicating risks.

Most research into general uncertainty communication seems to focus specifically on how we communicate risks. Risk and uncertainty are not perfect synonyms of one another. @Spiegelhalter2017 makes the distinction clear by specifying *risk* as *a precise random event* (such as a coin flip) while *uncertainty* (also sometimes referred to as abiguity) is imprecision (such as the estimated probability associated with a coin that might be biased). For this reason, risk is often seen as a specific case of uncertainty and that well defined sub area has a lot of detailled research. Some markers of probability that have common uses, probabilities such as 0 (impossible), 0.5 (a fair coin flip), 1 (certain) are easy for people to have an internal sense of *hullman paper*. It is incredibly hard to communicate small risks, because people cannot differentiate between 1/100 an 1/1000, and communicating low probability but high impact risks are particularly challenging [@Spiegelhalter2017]. This issue can be somewhat alleviated by communicating the likelihood of an event in relation to another event on a similar scale (such as being struck by lightning) [@Spiegelhalter2017]. The framing of how we communicate risks matters due to the affect heuristic, so there will be a difference in peoples behaviours if, for example, you tell people they have a 90% chance of survival vs a 10% chance of death, and when expressing a frequency, a larger numerator communicates a larger risk [@Spiegelhalter2017]. The reference class of a particular risk should be explicitly stated to avoid confusion, for example if we are communicating the chance of rain, the time period over which we are expecting this chance of rain (i.e. 50% chance today or this week) needs to be specified [@Spiegelhalter2017]. The problem with a risk focus is that it is only a small piece of the uncertainty pie, and very few of these findings can be reasonably assumed to go beyond risk communication. Even less can be assumed to translate to uncertainty visualisation.

@@

::: aside
This section still needs some citations from papers that communicate uncertainty, but the are *hard* to organise because of the ambiguous uncertainty definition problem. So they will all be communicating something different but you cant find out what it is until you are knee deep in the method section.
:::

@@

What is meant by "uncertainty" may seem obvious to some, but when you attempt to quantify or visualise it you will quickly find yourself asking, "uncertainty about... what?". Do you mean uncertainty on an estimate? On a forecast? How many steps ahead is this forecast? Are we only considering the uncertainty in the estimate or in the parameters or are we considering the possibility of measurement error or biased inputs? Signal and noise can only be untangled in the presence of a motivating question.

@@

## 4.1 Similar work

::: aside
Again, this is why I had that table, because these papers are impossible to organise without some metrics. Things that are highly important for uncertainty as I understand it, are not mentioned at all or only mentioned knee deep in the method, so reading and organising these papers such that the information can be combined is a bit of a nightmare. If I *dont* find this information uncertainty, I will end up with the same problem as *all* the other uncertainty visualisation literature reviews, which is that the information is *all conflicting and it is impossible to identify any results*. The other literature reviews *clearly state this*.
:::

There are a wealth of uncertainty communication and visualisation literature reviews that already exist.

@@

@@
